\vspace{\baselineskip} Approximate Bayesian Computation of the population of exoplanets \vspace{-2\baselineskip}

### 2016-10-05 paper/ms.tex
step
words about abc

\include{vc}

@@

@@

@@

\section{Toy problem: a Poisson process}

To demonstrate the issues with this method, let's start with the simplest toy
problem: a single observation of a Poisson process.
In this case, the correct posterior inference can be performed because the
correct likelihood is analytically tractable.
The generative model is
\begin{eqnarray}
N_\obs &\sim& p(N_\obs\mid\mu) = \frac{\mu^{N_\obs}\,e^{-\mu}}{N_\obs!}
\end{eqnarray}
with a single parameter $\mu$.
Placing a uniform prior on $\log\mu$ between $\mu_\mathrm{min}$ and
$\mu_\mathrm{max}$, the posterior probability becomes
\begin{eqnarray}
p(\mu\mid N_\obs) \propto \left\{\begin{array}{ll}
    \frac{1}{\mu}\frac{\mu^{N_\obs}\,e^{-\mu}}{N_\obs!}\quad & \quad
       \mathrm{if}\,\mu_\mathrm{min} \le \mu \le \mu_\mathrm{max} \\
    0 \quad & \quad \mathrm{otherwise}
\end{array}\right. \quad.
\end{eqnarray}

For this simple example, \dfmfig{abctoy} compares the inferences made using
\eq{heuristic-likelihood} to the inferences made using \abc\ with a
pseudo-likelihood
\begin{eqnarray}
p(N_\obs\mid\mu) &=& \int K(N_\simu,\,N_\obs;\,\epsilon)\,
    p(N_\simu\mid\mu) \dd N_\simu
\end{eqnarray}
where
\begin{eqnarray}
p(N_\simu\mid\mu) &=& \frac{\mu^{N_\simu}\,e^{-\mu}}{N_\simu!}
\end{eqnarray}
is the generative model.
For this demonstration, we use a Gaussian kernel
\begin{eqnarray}
K(N_\simu,\,N_\obs;\,\epsilon) \propto \exp\left(
    -\frac{(N_\simu - N_\obs)^2}{2\,\epsilon^2}
\right)
\end{eqnarray}
with $\epsilon \equiv 10^{-3}$.
The pseudo-likelihood is computed numerically as
\begin{eqnarray}
p(N_\obs\mid\mu) &\approx& \frac{1}{S}\sum_{s=1}^S
    K({N_\simu}^{(s)},\,N_\obs;\,\epsilon)
\end{eqnarray}
for ${N_\simu}^{(s)} \sim \mathrm{Pois}(\mu)$ and $S = 50$.

\begin{figure*}[p]~\\
\begin{center}
\includegraphics{figures/abctoy.pdf}
\end{center}
\caption{%
A toy example.
\dfmfiglabel{abctoy}}
\end{figure*}

>>>>

\include{vc}
\include{figures/abctoy}

@@

\section{A motivating example: inference with a Poisson process}

To demonstrate the issues with this method, let's start with the simplest toy
problem: a single observation of a Poisson process.
In this case, the correct posterior inference can be performed because the
correct likelihood is analytically tractable.
The generative model is
\begin{eqnarray}
N_\obs &\sim& p(N_\obs\mid\mu) = \frac{\mu^{N_\obs}\,e^{-\mu}}{N_\obs!}
\end{eqnarray}
with a single parameter $\mu$.
Placing a uniform prior on $\log\mu$ between $\mu_\mathrm{min}$ and
$\mu_\mathrm{max}$, the posterior probability becomes
\begin{eqnarray}\eqlabel{exact-posterior}
p(\mu\mid N_\obs) \propto \left\{\begin{array}{ll}
    \frac{1}{\mu}\frac{\mu^{N_\obs}\,e^{-\mu}}{N_\obs!}\quad & \quad
       \mathrm{if}\,\mu_\mathrm{min} \le \mu \le \mu_\mathrm{max} \\
    0 \quad & \quad \mathrm{otherwise}
\end{array}\right. \quad.
\end{eqnarray}

To test the methods, we generated a simulated ``data set'' of $N_\obs =
\abctoynobs$ from a Poisson process with $\mu_\mathrm{true} = \abctoytruth$.
Using this simulated data, \dfmfig{abctoy} compares the inferences made using
\eq{heuristic-likelihood} to the exact posterior inference using
\eq{exact-posterior} and to inferences made using \abc\ with a
pseudo-likelihood (the details will be given in the following section)
\begin{eqnarray}\eqlabel{pseudo-likelihood}
p(N_\obs\mid\mu) &=& \int K(N_\simu,\,N_\obs;\,\epsilon)\,
    p(N_\simu\mid\mu) \dd N_\simu
\end{eqnarray}
where
\begin{eqnarray}
p(N_\simu\mid\mu) &=& \frac{\mu^{N_\simu}\,e^{-\mu}}{N_\simu!}
\end{eqnarray}
is the generative model.
For this demonstration, we use a Gaussian kernel
\begin{eqnarray}
K(N_\simu,\,N_\obs;\,\epsilon) \propto \exp\left(
    -\frac{(N_\simu - N_\obs)^2}{2\,\epsilon^2}
\right)
\end{eqnarray}
with $\epsilon \equiv \abctoyeps$.
The pseudo-likelihood is computed numerically as
\begin{eqnarray}
p(N_\obs\mid\mu) &\approx& \frac{1}{S}\sum_{s=1}^S
    K({N_\simu}^{(s)},\,N_\obs;\,\epsilon)
\end{eqnarray}
for ${N_\simu}^{(s)} \sim \mathrm{Pois}(\mu)$ and $S = \abctoyS$.

From \dfmfig{abctoy}, it is clear that the \abc\ inference correctly recovers
the exact posterior while the heuristic method underestimates the mean of the
posterior and overestimates the variance.
In practice, this inconsistency increases when the true underlying
distribution is no longer Poisson.

The main point of this demonstration is that the heuristic method commonly
used to do inference with stochastic population models is not sufficient if we
want to do rigorous quantitative population inference.
Instead, \abc\ can be used to capture the exact posterior.

\begin{figure*}[p]~\\
\begin{center}
\includegraphics{figures/abctoy.pdf}
\end{center}
\caption{%
A comparison between a commonly used heuristic method for doing inference with
stochastic models and \abc.
The blue histogram shows the posterior inference for the rate parameter $\mu$
using the heuristic likelihood function from \eq{heuristic-likelihood}, the
orange histogram shows the \abc\ inference using the pseudo-likelihood from
\eq{pseudo-likelihood}, and the black curve shows the exact posterior from
\eq{exact-posterior}.
The vertical gray line shows the true value of $\mu$ that was used to generate
the data.
\dfmfiglabel{abctoy}}
\end{figure*}

@@

The problem statement for \abc\ is that you have a method of (stochastically)
simulating representative data from the model given a set of parameters but it
is computationally intractable to compute the likelihood of a given data set
conditioned on these same parameters.
In other words, you can sample $D_\simu \sim p(D_\simu\mid\theta)$ but you
can't directly evaluate $p(D_\obs\mid\theta)$ for any values $D_\obs$ and
$\theta$.
This means that you can't use standard \mcmc\ methods to draw samples from the
posterior probability $p(\theta\mid D_\obs) \propto
p(D_\obs\mid\theta)\,p(\theta)$.

To motivate \abc, let's start with an extreme method.
Let's say that sample a set of parameters $\theta \sim p(\theta)$ from your
prior and then simulate some data $D_\simu$ that just happens to
\emph{exactly} match the observed data $D_\obs$.
In this case, under your simulation model, this value of $\theta$ is a valid
sample from the posterior probability density $p(\theta\mid D_\obs)$ even
though you never evaluated the likelihood.
Now if you had an infinite amount of computing power and could draw a huge
number of samples following this procedure and rejecting all samples that
don't result in an exact match for your data then you could produce a
posterior sampling for $p(\theta\mid D_\obs)$ without ever evaluating the
likelihood.
Of course this method is absurd and intractable so \abc\ builds on this using
two observations
\begin{enumerate}
\item This method will still be correct if sufficient statistics of the
simulated data match the same sufficient statistics on the observed data.
\item The comparison between the simulation and the observation can be
``softened'' without significant loss of information.
\end{enumerate}
The first point is correct by definition but we will return to the relevant
subtleties below.
The second point is less obvious.

@@
